% Created 2017-11-09 jue 07:22
% Intended LaTeX compiler: pdflatex
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{color}
\usepackage{listings}
\usepackage{mathpazo}
\usepackage[usenames,svgnames,dvipsnames]{xcolor}
\hypersetup{colorlinks=true, linkcolor=Blue, urlcolor=Blue}
\author{Oscar Perpi침치n Lamigueiro}
\date{November 2017}
\title{Outline of Changes\\\medskip
\large Second Edition of ``Displaying Time Series, Spatial, and Space-Time Data with R''}
\hypersetup{
 pdfauthor={Oscar Perpi침치n Lamigueiro},
 pdftitle={Outline of Changes},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 25.2.2 (Org mode 9.1.2)}, 
 pdflang={English}}
\begin{document}

\maketitle


\section*{Time Series}
\label{sec:orgbb04cc8}

\begin{itemize}
\item Code have been revised, cleaned and improved. Bugs have been fixed.

\item After data creation or transformation, a text frame with the output of \texttt{summary()} has been included.

\item Code using \texttt{base::reshape} has been replaced by \texttt{reshape2::melt} and \texttt{reshape2::dcast}.
\end{itemize}

\subsection*{Time on the Horizontal Axis}
\label{sec:orgc386f10}

\begin{itemize}
\item Alternative code based on \texttt{ggplot2} for the Figure 3.9 (\emph{Level plot of differences between a daily temperature time \ldots{}}).

\item New ``Interactive Graphics'' section (Section 3.4):
\begin{itemize}
\item Some figures have an interactive alternative with code based on the packages \texttt{dygraphs}, \texttt{highcharter}, \texttt{plotly}, and \texttt{streamgraph}.
\item The \texttt{gridSVG} section has been revised and moved here (Section 3.4.4).
\item This section includes snapshots of the interactive graphics that will be available in the updated webpage.
\end{itemize}
\end{itemize}

\subsection*{Time as a Conditioning or Grouping Variable}
\label{sec:orgf3b4bac}

\begin{itemize}
\item Alternative code based on \texttt{ggplot2} (using package \texttt{GGally}) for the Figure 4.1 (\emph{Scatter plot matrix of the collection\ldots{}}).
\end{itemize}

\subsection*{Time as a Complementary Variable}
\label{sec:org4420047}

\begin{itemize}
\item New ``Interactive Graphics: animation'' section (Section 5.6):
\begin{itemize}
\item A new section based on the package \texttt{plotly} has been created (section 5.6.1).
\item Original sections using \texttt{googleVis} and \texttt{gridSVG} have been revised and moved here.
\item This section includes snapshots of the interactive graphics that will be available in the updated webpage.
\end{itemize}
\end{itemize}

\subsection*{About the Data}
\label{sec:orgf0f01e5}
\begin{itemize}
\item Data URLs have been checked and updated where needed.
\end{itemize}

\section*{Spatial Data}
\label{sec:org9a5e11b}

To Do.
\section*{Space-Time Data}
\label{sec:org8905af4}
To Do.
\end{document}